---
title: "Permutations and combinations"
slug: "permcomb"
author: "Mark Andrews"
date: 2019-01-30
tags: ["permutations", "combinations", "maths", "stats", "R"]
stub: false
description: |
  Permutations and combinations frequently arise in probability
  calculations. Here, we provide a brief introduction and 
  show how to calculate them in R.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(magrittr)

```


Permutations and combinations frequently arise in probability
calculations. For example, to calculate the probability of winning the
lottery, we would have to first calculate the number of possible
outcomes of a single lottery draw.

A *permutation* is a re-ordering of a set of items. If we had a set of
three items such as tea, coffee and sugar, for example, we could order
these three items in six ways, i.e. {tea, coffee, sugar}, {tea, sugar,
coffee}, {coffee, tea, sugar}, {coffee, sugar, tea}, {sugar, tea,
coffee} and {sugar, coffee, tea}. More generally, if we had $N$ items we
could order them in $N!$ ways or $N
\times (N-1) \times (N-2) \times \ldots 2 \times 1$. This is because if
you had $N$ items, there are $N$ possibilities for the first item in the
ordering, there are $N-1$ possibilities for the second, and so on.

We can also talk about the permutations of $N$ items taken $n$ at a
time. For example, let's say we have a class with 28 students. One desk
in the classroom has six seats. How many ways are there for how we could
seat different groups of students at this desk? Any one of the 28
students could sit on the first chair. Then any one of 27 students could
sit on the second chair, and so on until we have 23 possible students
left who could sit on the sixth and final chair. In general, the number
of permutations of $N$ items taken $n$ at a time is $$\begin{aligned}
\mathrm{Perm}(N,n) &= N \times (N-1) \times \ldots \times (N-n+1),\\
&= \frac{N \times (N-1) \times (N-2) \times \ldots 2 \times 1}{ (N-n) \times (N-n-1) \times (N-n-2) \times \ldots 2 \times 1},\\
&= \frac{N!}{(N-n)!}.\end{aligned}$$

In contrast to a permutation, a *combination* is a way of selecting a
subset of items from a larger set, but where the ordering of the subset
is irrelevant. For example, if we have a set of $N$ items, how many ways
could we choose $n$ items from this set (assuming $n<N$). This number is
related to $\mathrm{Perm}(N,n)$. In the example above, we have seen that
there are $28
\times 27 \times \ldots \times 23$ ways that we could seat 28 students
at a desk with 6 seats. This takes into account the order in which each
possible group of six students are seated. If do not care who sits
where, but just whether a given student is seated at the desk or not,
then we can divide the number of permutations by the number of ways of
ordering each group of 6 students. This is
$6 \times 5 \times 4 \times 3 \times 2 \times 1$ or $6!$. In other
words, the number of ways we could select a subset of 6 students from a
group of 28 is $$\mathrm{Comb}(28,6) = \frac{28 \times 27 \times
\ldots \times 23}{6 \times 5 \times 4 \times 3 \times 2 \times 1},$$ and
in general, $$\begin{aligned}
\mathrm{Comb}(N,n) &= \frac{\mathrm{Perm}(N,n)}{n!},\\
&= \frac{N!}{n!(N-n)!}.\end{aligned}$$ The term $\mathrm{Comb}(N,n)$ is
very commonly written as $\tbinom{N}{n}$, and read as "N choose n".

Here is a simple example of combinations. How many ways could be select
2 items from the set of tea, coffee and sugar? Let's first list the
possible permutations of the three items, taken two at a time: {tea,
coffee}, {coffee, tea}, {tea, sugar}, {sugar, tea}, {sugar, coffee},
{coffee, sugar}. There are six permutations, which is
$\mathrm{Perm}(3,2)$. However, clearly the sets {tea,coffee} and
{coffee,tea}, {tea, sugar} and {sugar, tea}, {sugar, coffee} and
{coffee, sugar} contain the same items, just in different orders. If we
ignore these orders, we see that there are only three possible subsets
of tea, coffee and sugar when we take two at a time. To get this, we
divided the number of permutations, i.e. $\mathrm{Perm}(3,2)=6$, by the
number of ways you could re-order the subsets of two items, i.e.
$2 \times
1 = 2$.

```{r, echo=F}
# If you change these, the LaTeX below does not adapt
N <- 59
m <- 6
```

Finally, consider the number of ways of winning the jackpot in the UK's
lotto. There are `r N` possible numbers, of which you choose `r m`. The draw
consists of drawing `r m` numbers randomly from the set of `r N`. If you match
the numbers drawn *in any order*, you will win or share in the jackpot.
What is the probability of winning this jackpot? To calculate this,
first we count the number of possible draws of `r m` numbers from the drum
containing the `r N` numbers. The first number drawn can be any one of `r N`
possible numbers, the second any one of `r N-1`, and so on until the sixth
which is any one of `r N-(m-1)` possible numbers. In other words, the total
numbers of draws is
$$`r N` \times `r N-1` \times `r N-2` \times  `r N-3` \times  `r N-4` \times  `r N-5` = 
`r exp(lfactorial(N) - lfactorial(N-6)) %>% format(big.mark=',') `.$$
This number takes into account the number of
different orders in which the six numbers could occur. In other words,
it counts the numbers $\{1,2,3,4,5,6\}$ as distinct from
$\{2,1,3,4,5,6\}$, and so on. The number of subsets of `r m` numbers taken
from `r N`, where their ordering is irrelevant, is the number listed above
divided by the number of ways we can re-order a group of six numbers,
which is $`r m`!$, i.e.
$$\frac{`r N` \times `r N-1` \times `r N-2` \times  `r N-3` \times  `r N-4` \times  `r N-5`}
{6 \times 5 \times 4 \times 3 \times 2 \times 1} 
= \frac{`r exp(lfactorial(N) - lfactorial(N-6)) %>% format(big.mark=',') `}
{720} = `r lchoose(N, m) %>% exp() %>% format(big.mark=',')`.$$
In other words, there are just 45 million different possible sets
of lottery draws and so the chance that your choice of numbers will win
is about 1 in 45 million.


## Permutations and combinations using R

Permutations and combinations ultimately just involve products and ratios of factorials. 
In R, we can calculate the factorial of any number using `factorial`. For example,
```{r}
x <- c(5, 3, 11) # just arbitrarily chosen numbers
factorial(x)     # their factorials
```
The problem is that factorials can get *huge* very quickly, and this can lead 
to [overflow](https://en.wikipedia.org/wiki/Integer_overflow). We can 
try to avoid this by using [logarithms](wiki/logarithms) of the factorials. For example,
```{r}
lfactorial(x)     # the logarithms of the factorials of x
```
In fact, R provides logarithsm of the $\mathrm{Comb}(N,n)$ function above as follows:
```{r}
lchoose(12, 3)    # the logarithm of 12 choose 3
```
We can then define our own permutations and combination functions:
```{r}
permutation = function(N, m, as_log=F) {
  log_perm <- lfactorial(N) - lfactorial(N-m)
  ifelse(as_log, log_perm, exp(log_perm))
}

combination = function(N, m, as_log=F) {
  log_comb <- lchoose(N, m)
  ifelse(as_log, log_comb, exp(log_comb))
}
```
We'll test this out with some of the calculations above:
```{r}
permutation(59, 6)
```
and
```{r}
combination(59, 6)
```
If you, like me, need commas to help read any numbers with around 6 or more numbers you can do the following:
```{r}
permutation(59, 6) %>% format(big.mark=',')
```
and
```{r}
combination(59, 6) %>% format(big.mark=',')
```


